---
title: 'Lesson 7: Advanced Raster Analysis'
author: "Ben Devries, Jan Verbesselt, Loïc Dutrieux"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
    number_sections: true
---

# Today's items

## ToDo in the morning

Go through the whole lesson and try to answer the questions below. We will address the questions in the lesson and discuss.

## Objective of today

Advanced Raster Analysis!

## Learning outcomes of today:

- carry out a supervised classification (random forest) on a series of raster layers
- construct a raster sieve using the `clump()` function
- work with thematic rasters

# Advanced Raster Analysis

## Landsat data analysis

Since being released to the public, the Landsat data archive has become an invaluable tool for environmental monitoring. With a historical archive reaching back to the 1970's, the release of these data has resulted in a spur of time series based methods. In this tutorial, we will work with time series data from the Landsat 7 Enhanced Thematic Mapper (ETM+) sensor. Landsat scenes are delivered via the USGS as a number of image layers representing the different bands captured by the sensors. In the case of the Landsat 7 Enhanced Thematic Mapper (ETM+) sensor, the bands are shown in Figure X. Using different combination of these bands can be useful in describing land features and change processes.

`insert figure here`

Part of a landsat scene, including bands 2-4 are included as part of the rasta package. These data have been processed using the LEDAPS framework (http://ledaps.nascom.nasa.gov/), so the values contained in this dataset represent surface reflectance, scaled by 10000 (ie. divide by 10000 to get a reflectance value between 0 and 1).

We will begin exploring these data simply by visualizing them.

```{r, message=FALSE, include=FALSE}
## load data
load("data/GewataB2.rda")
load("data/GewataB3.rda")
load("data/GewataB4.rda")

# check out the attributes
GewataB2
# some basic statistics using cellStats()
cellStats(GewataB2, stat=max)
cellStats(GewataB2, stat=mean)
# This is equivalent to:
maxValue(GewataB2)
# what is the maximum value of all three bands?
max(c(maxValue(GewataB2), maxValue(GewataB3), maxValue(GewataB4)))
# summary() is useful function for a quick overview
summary(GewataB2)

# put the 3 bands into a rasterBrick object to summarize together
gewata <- brick(GewataB2, GewataB3, GewataB4)
# 3 histograms in one window (automatic, if a rasterBrick is supplied)
hist(gewata)
```


When we plot the histogram of the rasterBrick, the scales of the axes and the bin sizes are not equivalent, which could be problematic. This can be solved by adjusting these paramters in `hist()`, which would take some extra work and consideration. Alternatively, the `rasterVis` package has enhanced plotting capabilities which make it easier to make more attractive plots with these considerations in mind.

```{r}
library(rasterVis)
# view all histograms together with rasterVis
histogram(gewata)
```

The `rasterVis` package has several other raster plotting types inherited from the lattice package. For multispectral data, one plot type which is particularly useful for data exploration is the scatterplot matrix, called by the `splom()` function.

```{r}
splom(gewata)
```

Calling `splom()` on a rasterBrick reveals potential correlations between the layers themselves. In the case of bands 2-4 of the gewata subset, we can see that band 2 and 3 (in the visual part of the EM spectrum) are highly correlated, while band 4 contains significant non-redundant information. 

**Given what we know about the location of these bands along the EM spectrum, how could these scatterplots be explained?** ETM+ band 4 (nearly equivalent to band 5 in the Landsat 8 OLI sensor) is situated in the near infrared (NIR) region of the EM spectrum and is often used to described vegetation-related features.


We observe a strong correlation between two of the Landsat bands of the gewata subset, but a very different distribution of values in band 4 (NIR). This distribution stems from the fact that vegetation reflects very highly in the NIR range, compared to the visual range of the EM spectrum. A commonly used metric for assessing vegetation dynamics, the normalized difference vegetation index (NDVI), explained in the previous lesson, takes advantage of this fact and is computed from Landsat bands 3 (visible red) and 4 (near infra-red).

In the previous lesson, we explored several ways to calculate NDVI, using direct raster algebra, `calc()` or `overlay()`. Since we will be using NDVI again later in this tutorial, let's calculated it again and store it in our workspace using `overlay()`.

```{r}
ndvi <- overlay(GewataB4, GewataB3, fun=function(x,y){(x-y)/(x+y)})
```

Aside from the advantages of `calc()` and `overlay()` regarding memory usage, an additional advantage of these functions is the fact that the result can be written immediately to file by including the `filename = "..."` argument, which will allow you to write your results to file immediately, after which you can reload in subsequent sessions without having to repeat your analysis.

## Classifying raster data

One of the most important tasks in analysis of remote sensing image analysis is image classification. In classifying the image, we take the information contained in the various bands (possibly including other synthetic bands such as NDVI or principle components). In this tutorial we will explore two approaches for image classification: 
  - supervised (random forest) classification
  - unsupervised (k-means).

### Supervised classification: Random Forest

The Random Forest classification algorithm is an ensemble learning method that is used for both classification and regression. In our case, we will use the method for classification purposes. Here, the random forest method takes random subsets from a training dataset and constructs classification trees using each of these subsets. Trees consist of *branches* and *leaves*. 

Branches represent nodes of the decision trees, which are often thresholds defined for the measured (known) variables in the dataset. Leaves are the class labels assigned at the termini of the trees. Sampling many subsets at random will result in many trees being built. Classes are then assigned based on classes assigned by all of these trees based on a majority rule, as if each class assigned by a decision tree were considered to be a *vote*. 

Figure [1](fig:random) gives a simple demonstration of how the random forest method works in principle. For a more complete description of the Random Forests classification method, see [here](http://stat-www.berkeley.edu/users/breiman/RandomForests/cc\textunderscore home.htm). For a ``friendlier" introduction, see this [presentation](http://www.slideshare.net/0xdata/jan-vitek-distributedrandomforest522013).


```{r random, echo=FALSE, fig.width=5, fig.align='center', fig.cap='Fig 1. Schematic showing how the Random Forest method constructs classification trees from random subsets of a training dataset. Each tree determines the labels assigned based on the training dataset. Once all trees are assembled, classes are assigned to unknown pixels based on the class which receives the majority of votes based on all the decision trees constructed.'}
library(png)
library(grid)
img <- readPNG("figs//randomForestDescription.png")
grid.raster(img)
```

One major advantage of the Random Forest method is the fact that an *Out of the Bag* (OOB) error estimate and an estimate of variable performance are performed. For each classification tree assembled, a fraction of the training data are left out and used to compute the error for each tree by predicting the class associated with that value and comparing with the already known class. This process results in a confusion matrix, which we will explore in our analysis. In addition an importance score is computed for each variable in two forms: the mean decrease in accuracy for each variable, and the Gini impurity criterion, which will also be explored in our analysis.

We should first prepare the data on which the classification will be done. So far, we have prepared three bands from a ETM+ image in 2001 (bands 2, 3 and 4) as a rasterBrick, and have also calculated NDVI. In addition, there is a Vegetation Continuous Field (VCF) product available for the same period (2000).

For more information on the Landsat VCF product, see (http://glcf.umd.edu/data/landsatTreecover/). This product is also based on Landsat ETM+ data, and represents an estimate of tree cover (in %). Since this layer could also be useful in classifying land cover types, we will also include it as a potential covariate in the random forest classification.

```{r, fig.align='center'}
# load the data and check it out
# @ can we show a bit more about this data set - where does it come from
# @ can we provide an option to download it themselve for any location?
library(raster)
load("data/vcfGewata.rda")
vcfGewata
plot(vcfGewata)
summary(vcfGewata)
hist(vcfGewata)
```

Note that in the `vcfGewata` rasterLayer there are some values much greater than 100, which are flags for water, cloud or cloud shadow pixels. To avoid these layers, we can assign a value of `NA` to these pixels so they are not used in the classification.

```{r}
vcfGewata[vcfGewata > 100] <- NA
# look at revised summary stats
summary(vcfGewata)
# plot(vcfGewata)
# hist(vcfGewata)
```

To perform the classification in R, it is best to assemble all covariate layers (ie. those layers contaning predictor variable values) into one rasterBrick object. In this case, we can simply append these new layers (NDVI and VCF) to our existing rasterBrick (currently consisting of bands 2, 3, and 4). But first, let's rescale the NDVI layer by 10000 (just as the reflectance bands 2, 3, and 4 have been scaled) and store it as an integer raster. In Lesson 5 we encountered different data types for rasters, so we will not go into more detail here.

```{r, eval=TRUE, fig.align='center'}

# multiply all values by 10000
ndvi <- calc(ndvi, fun = function(x) floor(x*10000))
# change the data type
# see ?dataType for more info
dataType(ndvi) <- "INT2U"
# name this layer to make plots interpretable
names(ndvi) <- "NDVI"
# make the covariate rasterBrick
covs <- addLayer(gewata, ndvi, vcfGewata)
plot(covs)
```

<!---

<<covariants, fig=FALSE>>=

@

For this exercise, we will do a very simple classification for 2001 using three classes: forest, cropland and wetland. While for other purposes it is usually better to define more classes (and possibly fuse classes later), a simple classification like this one could be useful, for example, to construct a forest mask for the year 2001.

<<trainingPoly, fig=FALSE, results=hide>>=
# load the training polygons
data(trainingPoly)
# superimpose training polygons onto ndvi plot
plot(ndvi)
plot(trainingPoly, add = TRUE)
@

The training classes are labelled as string labels. For this exercise, we will need to work with integer classes, so we will need to first 'relabel' our training classes. There are several approaches that could be used to convert these classes to integer codes. In this case, we will first make a function that will reclassify the character strings representing land cover classes into integers based on the existing factor levels.

<<reclassify>>=
# inspect the data slot of the trainingPoly object
trainingPoly@data
# the 'Class' column is actually an ordered factor type
trainingPoly@data$Class
str(trainingPoly@data$Class)

# define a reclassification function which substitutes
# the character label for the factor level (between 1 and 3)
reclass <- function(x){
  which(x==levels(trainingPoly@data$Class))
}

# use sapply() to apply this function over each element of the 'Class' column
# and assign to a new column called 'Code'
trainingPoly@data$Code <- sapply(trainingPoly@data$Class, FUN=reclass)
@

To train the raster data, we need to convert our training data to the same type using the \code{rasterize()} function. This function takes a spatial object (in this case a polygon object) and transfers the values to raster cells defined by a raster object. Here, we will define a new raster containing those values.

<<training_plot, fig=FALSE, results=hide>>=
# assign 'Code' values to raster cells (where they overlap)
classes <- rasterize(trainingPoly, ndvi, field='Code')
# set the dataType of the raster to INT1U
# see ?dataType for more information
dataType(classes) <- "INT1U"
# define a colour scale for the classes
# corresponding to: cropland, forest, wetland
cols <- c("orange", "dark green", "light blue")
# plot without a legend
plot(classes, col=cols, legend=FALSE)
# add a customized legend
legend("topright", legend=c("cropland", "forest", "wetland"), fill=cols, bg="white")
@

(Note: there is a handy ``\code{progress="text"}" argument, which can be passed to many of the raster package functions and can help to monitor processing. Try passing this argument to the \code{rasterize()} command above).

Our goal in preprocessing these data is to have a table of values representing all layers (covariates) with \emph{known} values/classes. To do this, we will first need to create a version of our rasterBrick only representing the training pixels. Here the \code{mask()} function from the raster package will be very useful.

<<masked_covariates, fig=FALSE>>=
covmasked <- mask(covs, classes)
plot(covmasked)
# add the classes layer to this new brick
names(classes) <- "class"
trainingbrick <- addLayer(covmasked, classes)
plot(trainingbrick)
# Note that in this plot, the 'class' legend is not meaningful
# This plot is useful simply to check the available layers
@

Now it's time to add all of these values to a data.frame representing all training data. This data.frame will be used as an input into the RandomForest classification function. We will use \code{getValues()} to extract all of the values from the layers of the rasterBrick.

<<valuetable, results=hide>>=
# extract all values into a matrix
valuetable <- getValues(trainingbrick)
# convert to a data.frame and inspect the first and last rows
valuetable <- as.data.frame(valuetable)
head(valuetable)
tail(valuetable)
@

In inspecting this training data.frame, you will notice that a significant number of rows has the value NA for the class column, which will be problematic during the training phase. The rows with class=NA represent pixels found outside the training polygons, and these rows should therefore be removed before going ahead with deriving the Random Forest model.

<<valuetable2, results=hide>>=
# keep only rows where valuetable$classes has a value
valuetable <- valuetable[!is.na(valuetable$class),]
head(valuetable)
tail(valuetable)

# convert values in the class column to factors
valuetable$class <- factor(valuetable$class, levels = c(1:3))
@

Now we have a convenient reference table which contains, for each of the three defined classes, all known values for all covariates. Before proceeding with the classification, let's visualize the distribution of some of these covariates using \code{ggplot()}. (Note: to make the following plots more readable, we will add a column with the class labels as characters. But we will not use this column when performing the classification.)

<<prep_ggplots, fig=FALSE>>=
# add a label column to valuetable
valuetable$label <- with(valuetable, ifelse(class==1, "cropland", 
                                            ifelse(class==2, "forest", "wetland")))
# see ?ifelse() for more information
@
<<NDVIVCFB3B4_ggplots, fig=FALSE, eval=FALSE>>=
# Now make the ggplots using valuetable$label to split the data into facets

# 1. NDVI
p1 <- ggplot(data=valuetable, aes(x=NDVI)) + 
  geom_histogram(binwidth=300) + 
  facet_wrap(~ label) +
  theme_bw()
p1

# 2. VCF
p2 <- ggplot(data=valuetable, aes(x=vcf2000Gewata)) +
  geom_histogram(binwidth=5) +
  labs(x="% Tree Cover") +
  facet_wrap(~ label) +
  theme_bw()
p2
# 4. Bands 3 and 4
p3 <- ggplot(data=valuetable, aes(x=gewataB3, y=gewataB4)) +
  stat_bin2d() +
  facet_wrap(~ label) +
  theme_bw()
p3

@
<<B2B3_ggplots, fig=TRUE, include=FALSE>>=
# 4. Bands 2 and 3
p4 <- ggplot(data = valuetable, aes(x=gewataB2, y=gewataB3)) +
  stat_bin2d() +
  facet_wrap(~ label) +
  theme_bw()
p4

@

\begin{figure}[!htp]
\centering
\includegraphics[width=\textwidth]{Lesson_6-B2B3_ggplots}
\caption{Band2-Band3 scatterplots for the three training classes.}
\label{B2B3_scatterplots}
\end{figure}

We can see from these distributions (e.g. Figure \ref{B2B3_scatterplots}) that these covariates may do well in classifying forest pixels, but we may expect some confusion between cropland and wetland (although the individual bands may help to separate these classes). When performing this classification on large datasets and with a large amount of training data, now may be a good time to save this table using the \code{write.csv()} command, in case something goes wrong after this point and you need to start over again.

Now it is time to build the Random Forest model using the training data contained in the table of values we just made. For this, we will use the "randomForest" package in R, which is an excellent resource for building such types of models. Using the \code{randomForest()} function, we will build a model based on a matrix of predictors or covariates (ie. the first 5 columns of valuetable) related to the response (the 'class' column of valuetable).

<<clean_valuetable>>=
# NA values are not permitted in the covariates/predictor columns
# keep only the rows with containing no NA's
valuetable <- na.omit(valuetable)
@
<<randomforest, eval=FALSE>>=
# construct a random forest model
# covariates (x) are found in columns 1 to 5 of valuetable
# training classes (y) are found in the 'class' column of valuetable
# caution: this step takes fairly long!
# but can be shortened by setting importance=FALSE
library(randomForest)
modelRF <- randomForest(x=valuetable[,c(1:5)], y=valuetable$class,
                        importance = TRUE)
@
<<RF_background, echo=FALSE, results=hide>>=
library(randomForest)
if(!file.exists("data/modelRF.rda")){
  modelRF <- randomForest(x = valuetable[,c(1:5)], y = valuetable$class,
                        importance = TRUE)
  save(modelRF, file='data/modelRF.rda', compress="bzip2", ascii=FALSE)
} else {
  load('data/modelRF.rda')
}
@

Since the random forest method involves the building and testing of many classification trees (the 'forest'), it is a computationally expensive step (and could take alot of memory for especially large training datasets). When this step is finished, it would be a good idea to save the resulting object with the \code{save()} command. Any R object can be saved as an .rda file and reloaded into future sessions.

The resulting object from the \code{randomForest()} function is a specialized object of class "randomForest", which is a large list-type object packed full of information about the model output. Elements of this object can be called and inspected like any list object.

<<inspect_RFmodel, results=hide>>=
# inspect the structure and element names of the resulting model
modelRF
class(modelRF)
str(modelRF)
names(modelRF)
# inspect the confusion matrix of the OOB error assessment
modelRF$confusion
# to make the confusion matrix more readable
colnames(modelRF$confusion) <- c("cropland", "forest", "wetland", "class.error")
rownames(modelRF$confusion) <- c("cropland", "forest", "wetland")
modelRF$confusion
@

Since we set 'importance=TRUE', we now also have information on the statistical importance of each of our covariates which we can visualize using the \code{varImpPlot()} command.

<<varImpPlot, fig=TRUE, include=FALSE>>=
varImpPlot(modelRF)
@

\begin{figure}[!htp]
\centering
\includegraphics[width=0.6\textwidth]{Lesson_6-varImpPlot}
\caption{Variable importance plots for a Random Forest model showing the mean decrease in accuracy (left) and the decrease in Gini Impurity Coefficient (right) for each variable.}
\label{varimp}
\end{figure}

These two plots give two different reports on variable importance (see \code{?importance()}). First, the mean decrease in accuracy indicates the amount by which the classification accuracy decreased based on the OOB assessment. Second, the Gini impurity coefficient gives a measure of class homogeneity. More specifically, the decrease in the Gini impurity coefficient when including a particular variable is shown in the plot \footnote{From Wikipedia: "Gini impurity is a measure of how often a randomly chosen element from the set would be incorrectly labeled if it were randomly labeled according to the distribution of labels in the subset". See \url{http://en.wikipedia.org/wiki/Decision_tree_learning}}.

In this case, it seems that Gewata bands 3 and 4 have the highest impact on accuracy, while bands 3 and 2 score highest with the Gini impurity criterion (Figure \ref{varimp}). For especially large datasets, it may be helpful to know this information, and leave out less important variables for subsequent runs of the \code{randomForest()} function.

Since the VCF layer included NA's (which have also been excluded in our results) and scores relatively low according to the mean accuracy decrease criterion, try to construct an alternate random forest model as above, but excluding this layer. What effect does this have on the overall accuracy of the results (hint: compare the confusion matrices of the original and new outputs). What affect does leaving this variable out have on the processing time (hint: use \code{system.time()})?

Now we can apply this model to the rest of the image and assign classes to all pixels. Note that for this step, the names of the raster layers in the input brick (here 'covs') must correspond to the column names of the training table. We will use the \code{predict()} function from the raster package to predict class values based on the random forest model we have just constructed. This function uses a pre-defined model to predict values of raster cells based on other raster layers. This model can be derived by a linear regression, for example. In our case, we will use the model provided by the \code{randomForest()} function we applied earlier.

<<predictRaster, fig=TRUE, include=FALSE, results=hide>>=
# check layer and column names
names(covs)
names(valuetable)
# predict land cover using the RF model
predLC <- predict(covs, model=modelRF, na.rm=TRUE)
# plot the results
# recall: 1 = cropland, 2 = forest, 3 = wetland
cols <- c("orange", "dark green", "light blue")
plot(predLC, col=cols, legend=FALSE)
legend("bottomright", legend=c("cropland", "forest", "wetland"), fill=cols, bg="white")
@
\begin{figure}[!htp]
\centering
\includegraphics[width=0.6\textwidth]{Lesson_6-predictRaster}
\caption{Resulting land cover map using a Random Forest classifier.}
\label{RandomForestLC}
\end{figure}

Note that the \code{predict()} function also takes arguments that can be passed to \code{writeRaster()} (eg. \code{filename = "..."}, so it would be a good idea to write to file as you perform this step (rather than keeping all output in memory).

-->

# Today's summary

We learned about:

- test
- test

# Excercise of today

## Optional for bonus points

# More info

About projections and code: https://www.nceas.ucsb.edu/scicomp/recipes/projections

# References